---
title: 'Lab 2: Graphing and Data Manipulation'
author: "Sergio Garcia-Rios"
date: "Feb 11 2019"
output: 
  html_document: 
    theme: simplex
    toc: yes
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
library(dplyr)
library(ggplot2)
library(gapminder)
```



The objectives of this lab are:


1. Manipulate data using the verbs in the **dplyr**
2. Use the pipe operator `%>%` to simplify complicated code by chaining expressions together
3. Create plots using the ggplot2 package.
*****

## Libraries and Data



For this lab we will use the following libraries `dplyr` and `ggplot2`. We can load those individually or use `tidyverse` which is not really a package but a collection of packages. Lately, I only load `tidyverse` since it contains all the packages that I frequently use. Remember, make sure you have installed these packages before you load them.



## Gapminder Data

You will be using the gapminder data again.

In the first lab, you loaded the data from a .csv file. In this lab, you will be using the same data, but as it is distributed in **gapminder** package.

To load a data set included with an R package, use the `data()` function.

You can see which data sets are included in a package. gapminder is not the only one most packages have some data set within the package. See also for instance `ggplot2`
```{r eval=FALSE}
data(package = "gapminder")
data(package = "ggplot2")

```


**Challenge**

- Which data set in the `gapminde` package is country data?
- Load that data using the `data` function
- Explore the dimensions of the data and it's variables


```




```



Load the gapminder data
```{r, echo = F}

data("gapminder")
head(gapminder)
```

## Introduction to `dplyr`

**dplyr** is a package for data manipulation. It provides a few core *verbs* and most data manipulations can be done by combining these verbs together --- something which becomes even easier with the `%>%` operator.

-  `filter()`: subset observations by logical conditions
-  `slice()`: subset observations by row numbers
-  `arrange()`: sort the data by variables
-  `select()`: select a subset of variables
-  `rename()`: rename variables
-  `distinct()`: keep only distinct rows
-  `mutate()` and `transmute()`: adds new variables
-  `group_by()`: group the data according to variables
-  `summarise()`: summarize multiple values into a single value
-  `sample_n()` and `sample_frac()`: select a random sample of rows

**dplyr** also offers the function `glimpse` to quickly view the data
```{r}
glimpse(gapminder)
```

### Exploring our Data

We are ready to begin exploring our data-set in more depth.

For this lab we want to explore the relationship between life expectancy and GDP.
Let's use some `dplyr` verbs to explore our data.
For you Stata users missing "if statements" let's begin with `filter()`

```{r results='hide'}
filter(gapminder, lifeExp < 29)
filter(gapminder, country == "Rwanda")
```

You can combine filter statements.


Including multiple logical statements is equivalent to combining them with "and".


This will give all observations in "Africa", before 1966, and which have a life expectancy less than 40.


```{r results='hide'}
filter(gapminder, continent == "Africa", year < 1966, lifeExp < 40)
```

That is equivalent to
```{r results='hide'}
filter(gapminder, continent == "Africa" & year < 1966 & lifeExp < 40 )
```


To combine logical statements with "or" you need to explicitly use `|`.
To find observations from Afghanistan or Albania,
```{r results='hide'}
filter(gapminder, country == "Afghanistan" | country == "Albania")
```


### Use `arrange` to sort columns in  a given order

Because the world is not always ordered the way we want it
```{r}
arrange(gapminder, pop) 

arrange(gapminder, -pop) 
```



### Use `select()` to subset the data on variables or columns.

Most of the times we don't need to see all the variables and are often interested in just a few of them. Here's a conventional call:
```{r}
select(gapminder, year, lifeExp) 
```


**Challenge**

Using a combination of `filter`, `select`, and `slice` to create data frames to show only year and life expectancy of Cambodia for the first two observations

```




```

```{r, echo=FALSE}
df1 <- filter(gapminder,country == "Cambodia")
df2 <- select(df1, year, lifeExp)
df3 <- slice(df2, 1:2)
```

### Use `%>%` to join the XXI century

Before we go any further, we should introduce  the pipe operator that `dplyr` imports from the **magrittr** package.

This is going to change your (data) analytical life.


Notice we can do the same computation as above but without having to create new objects. We are basically passing down the result from the previous line into the following. 

```{r, echo=F}
gapminder %>% 
  filter(country == "Cambodia") %>%
  select(year, lifeExp) %>%
  slice(1:2)

```

We think of the `%>%` operator as a _then_ statement. So in the previous line we:

- Called the gapminder data, _then_
- Filtered rows to those where country was Cambodia,  *then*
- We selected only `year` and `lifeExp` columns, _then_
- Sliced the data to see only the first two observations


### Use `mutate()` to add  new variables

Imagine we wanted to recover each country's GDP. We do have data for population and GDP per capita. what do we do?

-  Yes we multiply, let's create a new variable called GDP that brings back the gross amount

```{r, echo=FALSE}
gapminder <- 
  gapminder %>%
  mutate(gdp = pop * gdpPercap)

gapminder
```

So... GDP is almost useless because it doesn't give a base line and that is why we often use per capita, but a baseline is often more useful, how about comparing it to another country say, USA? 

Yes USA, USA, USA!

Let's  create first a data frame containing only US data, we use `filter` here. We are also need to change one of the variable name

```{r}
just_usa <- gapminder %>%
  filter(country == "United States") %>%
  select(year, gdpPercap) %>%
  rename(usa_gdpPercap = gdpPercap)
```

We can join (or merge) the data set to the gapminder data using the `left_join` function.


There are are several ways to merge data sets with **dplyr** (left join, right join, inner join, and outer join). They serve different purposes here we use `left_join()`


```{r}
gapminder <- 
  left_join(gapminder, just_usa, by = c("year")) 
```


No we can take `gdpPercap` and divide it by  `usa_gdpPercap to obtain a relative to the US figure.

```{r}
gapminder <- 
gapminder %>% 
  mutate(gdpPercapRel = gdpPercap / usa_gdpPercap)
```

Now, summarize the relative GDP
```{r}
gapminder %>%
  select(gdpPercapRel) %>%
  summary()

```

Nice, now we can do something like this:

Look at the GDP per capita  of Mexico and Canada relative to US by year
```{r}
gapminder %>%
  filter(country == "Canada") %>%
  select(country, year, gdpPercap, usa_gdpPercap, gdpPercapRel) 


gapminder %>%
  filter(country == "Mexico") %>%
  select(country, year, gdpPercap, usa_gdpPercap, gdpPercapRel) 

```


Or this:


```{r}
df_NAFTA<-
gapminder %>%
  filter(country %in% c("Mexico", "Canada")) %>%
  select(country, year, gdpPercap, usa_gdpPercap, gdpPercapRel) 

```



**Challenge**

What about life expectancy? Create a relative to life expectancy variable, compare the three NAFTA countries US, Canada and Mexico


```




```



```{r, echo = FALSE}

just_usa_life <- gapminder %>%
  filter(country == "United States") %>%
  select(year, lifeExp) %>%
  rename(usa_lifeExp = lifeExp)


gapminder <- 
  left_join(gapminder, just_usa_life, by = c("year")) 

gapminder <- 
gapminder %>% 
  mutate(lifeExpRel = lifeExp / usa_lifeExp)


gapminder %>%
  select(lifeExpRel) %>%
  summary()

```





## Plotting with ggplot2

For the most part, to visualize results in this course we will be using the graphics package **ggplot2**, which is one of the most popular, but it is only one of several graphics packages in R.[^1]

Unlike many other graphics systems, functions in **ggplot2** do not correspond to separate types of graphs.
There are not scatterplot, histogram, or line chart functions _per se_.
Instead plots are built up from component functions.

Main components:


1. Data
2. Aesthetics: Maps variables in the data to visual properties: position, color, size, shape, line type ...
3. Geometric objects: The specific shapes that are drawn: points, lines,


Other:

4. scales: How variables values map to "computer" values.
5. stat: summarize or transform the data. e.g. bin data and count in histogram; run a regression to get a line.
5. facet: create mini-plots of data subsets

Let's continue using the gapminder data, take another look at it
```{r}
glimpse(gapminder)
```

Great, let the plotting begin:

```{r eval=FALSE}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp))
```


This just initializes the plot with  the basic mapping. We still need to tell `ggplot` the geometric object (or `geoms`) that we will use to represent the data in this mapping.


Seems like points would be a good representation, let's use `geom_point`

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

That looks okay but it would probably look be better if we log transform.  Notice that we don't have to create a new variable we can just  do the transformation inside the aesthetics.

```{r}
ggplot(gapminder, aes(x = log10(gdpPercap), y = lifeExp)) +
  geom_point()
```


A better way to log transform
```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + 
  scale_x_log10()
```


Note the common workflow: gradually build up the plot you want,  re-define the plot by adding (literally) new elements.


Now, let's vary add another variable by having those points be represented by different colors according to the continent they belong. 

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  scale_x_log10()
```

Let's try address over-plotting: Set alpha transparency and size to a value. There is a even better way to do it but this is ok for now. 

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(alpha = .3, size = 3) +
  scale_x_log10()
```




Now, add a fitted curve or line, let's forget about continent for now

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = .3, size = 3) +
  geom_smooth() +
  scale_x_log10()
```

Let's remove the confidence band

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point() + 
  geom_smooth(lwd = 2, se = FALSE) + 
  scale_x_log10()
```

The default of `geom_smooth` follows a  local polynomial regression fitting (aka LOESS) but we can force any other type of fitting. Let's try a linear model `lm`

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point() + 
  geom_smooth(lwd = 2, se = FALSE,  method = "lm") + 
  scale_x_log10() 
```


That's great but I actually want to revive our interest in continents!

```{r}

ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) + 
  geom_point() + 
  geom_smooth(lwd = 2, se = FALSE,  method = "lm") + 
  scale_x_log10() 
```

That's a lot of info. Let's try something else called faceting. This creates various panels or subplots by a given variable.
```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) + 
  geom_point() + 
  scale_x_log10() + 
  facet_wrap(~ continent)

```
Still want best fit lines? Let's add them
```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) + 
  geom_point(alpha = .2) + 
  geom_smooth() + 
  scale_x_log10()  +
  facet_wrap(~ continent)
```


Notice what `color = continent` is doing here: it's affecting both the points and the line. Let's add some aesthetics specific to `geom_point` so that it changes only the points.


```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point(aes(color = continent, alpha = .2)) + 
  geom_smooth() + 
  scale_x_log10()  +
  facet_wrap(~ continent)
```




**Challenge**

* Plot lifeExp against year
* Make mini-plots, split out by continent
* Add a fitted smooth and/or linear regression, w/ or w/o faceting


```




```


***

Great! So we need to account for time. Let's use some of the cool capabilities of the extensions in the `ggplot`verse with the `gganimate` package


```{r, eval =F}
library(gganimate)

ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~ continent) +
  transition_time(year) +
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  theme_bw() 

```

Ok, back to the analysis. What if I am only interested in the US?
```{r}
ggplot(filter(gapminder, country == "United States"),
       aes(x = year, y = lifeExp)) +
  geom_line() +
  geom_point()

```


Let's just look at five countries

```{r}
some_countries <- c("United States", "Canada", "Rwanda", "Cambodia", "Mexico")

ggplot(filter(gapminder, country %in% some_countries),
       aes(x = year, y = lifeExp, color = country)) +
  geom_line() +
  geom_point()
```


So what's up with Mexico?

* NAFTA? Higher GDP?

Not really... Let's add yet another variable, GDP
```{r}
ggplot(subset(gapminder, country %in% some_countries),
       aes(x = year, y = lifeExp, color = country)) +
  geom_line() +
  geom_point(aes(size=gdpPercap))
```

You can change the way the plot looks overall using `theme`

```{r}
ggplot(subset(gapminder, country %in% some_countries),
       aes(x = year, y = lifeExp, color = country)) +
  geom_line() +
  geom_point(aes(size=gdpPercap)) +
  theme_minimal()
```

In addition to the themes included with ggplot, several other themes are available in the [ggthemes](http://cran.r-project.org/web/packages/ggthemes/index.html) package.

## References

- Hadley Wickham's tutorials from useR 2014: Video [Part I](https://www.youtube.com/watch?v=8SGif63VW6E), [Part II](https://www.youtube.com/watch?v=Ue08LVuk790), [Slides and Tutorial](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a).
- The vignettes for [dplyr](http://cran.rstudio.com/web/packages/dplyr/) and [tidyr](http://cran.rstudio.com/web/packages/dplyr/)


[^1]: You may encounter these other packages in other classes, or code samples online.

    - base graphics: included with R. Functions like `plot`, `barplot`, `hist`. See http://www.statmethods.net/graphs/index.html
    - lattice graphics: Functions including `barchart`, `densityplot`, `dotplot`, `xyplot`, `histogram`. See http://www.statmethods.net/advgraphs/trellis.html
    - - Other graphics packages designed for the web: [ggvis](http://ggvis.rstudio.com/), [rcharts](http://rcharts.io/), [plotly](https://plot.ly/)
    - - [tile](http://faculty.washington.edu/cadolph/?page=60) Chris Adolph's graphics package.

[^2]:

* * *
- Derived from Jennifer Bryan, "hello ggplot2!", https://github.com/jennybc/ggplot2-tutorial/blob/master/ggplot2-tutorial-slides.pdf. License: CC-BY-NC
- Derived from Karthik Ram, "A quick introduction to ggplot2", [Speakerdeck](http://inundata.org/2013/04/10/a-quick-introduction-to-ggplot2/), [github](https://github.com/karthik/ggplot-lecture). License: [CC-BY](http://creativecommons.org/licenses/by/2.0/)
-->


<!--  LocalWords:  'Lab Manipulation' html dplyr summarise frac gd df
 -->
<!--  LocalWords:  ggplot2 gapminder url gdf delim str tbl gapminder Stata
 -->
<!--  LocalWords:  lifeExp magrittr gdp gdpPercap usa gdpPercapRel gg
 -->
<!--  LocalWords:  NAFTA geoms aes colour ggplot log10 workflow 'p'
 -->
<!--  LocalWords:  lwd se lm Facetting gDat
 -->
