---
title: "Lab 2 - Intro to R and RStudio"
author: "Sergio Garcia-Rios"
date: "Feb 12 2018"
output:
  html_document:
    theme: cosmo
    highlight: haddock
---

```{r setup, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(gapminder)
```

* * *


Regular text

**Load Gapminderdata: **

This is how you load the data
```{r}
data(package = "gapminder")

data("gapminder")
```



```{r}
glimpse(gapminder)
```



```{r}
filter(gapminder, lifeExp <30, country == "Rwanda")


filter(gapminder, country == "Rwanda")


gapminder_low_lifeEXp <- 
  filter(gapminder, continent == "Africa", year < 1966, lifeExp < 40)


select(gapminder_low_lifeEXp, -continent)


gapminder2 <-
  gapminder %>%
  filter(continent == "Africa", year < 1966, lifeExp < 40) %>% 
  select(-continent)




```
Challenge

Using a combination of filter, select and and slice show only year and life expectancy of Cambodia for the first two observations


```{r}
gapminder %>% filter(country == "Cambodia") %>% 
  select(lifeExp, year) %>% 
  slice(2:3)

```






Now, imagine we wanted to recover each country’s GDP. We do have data for population and GDP per capita. what do we do?


```{r}
gapminder <-
gapminder %>% 
mutate(pop_1000 = pop/1000)


gapminder <-
gapminder %>% 
mutate(gdp = pop * gdpPercap)

```




```{r}
just_usa <- 
  gapminder %>% 
  filter(country == "United States") %>% 
  select(year, gdpPercap) %>% 
  rename(usa_gdpPercap = gdpPercap)
```



```{r}
gapminder <-
  left_join(gapminder, just_usa, by = c("year")) %>% 
  mutate(gdpPercapRel = gdpPercap / usa_gdpPercap)
```

```{r}
gapminder %>% 
  select(gdpPercapRel) %>% 
  summary()
```



```{r}
gapminder %>% 
  filter(country %in% c("United States", "Canada", "Mexico")) %>% 
  select(country, year, gdpPercap, usa_gdpPercap, gdpPercap)
```



```{r}
ggplot(gapminder, aes(x = log10(gdpPercap) ,y = lifeExp, color = continent)) +
  geom_smooth() +
  geom_jitter(alpha =.1) +
  theme_bw()+
  labs(title = "Title here", y= "Life Exp", x = "GDP per Cap")+
  facet_wrap(~ continent)


```









```{r}
ggplot(gapminder %>% filter(country== "United States"), aes(x =year, y = lifeExp, color =continent))+
  geom_smooth()+
  geom_jitter()

```


```{r}
some_countries <-gapminder %>% filter(country %in% c("United States", "Canada", "Rwanda", "Cambodia", "Mexico"))


ggplot(some_countries, aes(x = year, y = lifeExp, color = country)) +
  geom_line() +
  geom_point(aes(size=gdpPercap))+
  theme_bw()



```










#### Exercise 1: 

Now these are the exercises


```{r}
# enter your code for Exercise 1 here
```

#### Exercise 2:



```{r}
# enter your code for Exercise 2 here
```

#### Exercise 3:

```{r}
# enter your code for Exercise 3 here
```

* * *






