---
title: "Lab 4: Multivariate Linear Regression, and Predicted Values"
author: "Sergio Garcia-Rios"
output: 
  html_document: 
    theme: simplex
    toc: yes
    css: ../labs.css
editor_options: 
  chunk_output_type: console
---
```{r echo = FALSE, results = 'hide'}
knitr::opts_chunk$set(warning = FALSE, message = TRUE, eval = F)
```

For this lab we will use the replication [data]( https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/17976) from [Michael Ross](http://www.sscnet.ucla.edu/polisci/faculty/ross/)'s "The Oil Curse: How Petroleum Wealth Shapes the Development of Nations."

We will be exploring the relationship between oil dependency and democracy.

## Initial Setup


This lab will use some libraries you've seen before and we should load them now plus `broom`and `coefplot` which you might need to install

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(coefplot)
```

## Reading in the Data

Read in the Ross data

```{r, results='hide'}
rossdata <- read.csv("https://raw.githubusercontent.com/GarciaRios/govt_6029/master/data/ross_2012.csv",   stringsAsFactors = FALSE)

head (rossdata)
glimpse(rossdata)
```




This is a pretty big data-set, we do not need all the variables let's subset our data to include only the variables we will use.

1. *Using only what you need:*
    - Create a new data-set containing only:
      -  `cty`
      -  `year`
      -  `polity` 
      -  `logoil_gdpcap2000_sup_Q_1`
      -  `logGDPcap`
      -  `oecd`
    - Call this new data-frame `rossdata_short`

```{}



```

```{r, echo=FALSE}
rossdata_short <- rossdata %>% 
  select(cty, year, polity, logoil_gdpcap2000_sup_Q_1, logGDPcap, oecd)
```



## Data Management


Some of those names are too long, we should change it to something meaningful and short. This can be done easily using `dplyr` and `rename`.

```{r}
rossdata_short <- rossdata_short %>%
  rename(oil = logoil_gdpcap2000_sup_Q_1, 
         gdp = logGDPcap)
```

This data-frame is way easier to glimpse at:
```{r}
glimpse(rossdata_short)

```

A lot of missing values here. Let's omit them and then look at a summary of the data set
```{r}
rossdata_short <- rossdata_short %>% 
  na.omit()

rossdata_short %>%
  summary()
```

Finally! we are ready to start data-analyzing..

## Sacatterplots


We are going to be exploring the relationship between Democracy level (polity) and other covariates.

Let's explore these relationships with plots.

2. *Preliminary Exploration*
    -Create a plot that explores the relationship between democracy level and at least another variable then try to include more than two covariates using different colors and shapes.

```{}


```


Okay, let's do it together 

We begin simple... 

```{}



```

```{r, echo=FALSE}
ggplot(rossdata_short, aes(x = gdp, y = polity)) +
  geom_point() + 
  theme_bw() 

```

Unfortunately, a simple scatter plot makes it hard to detect any relationship. However, `ggplot2` makes it easy to add different colors and shapes which might help identify trends.

```{}



```

```{r, echo=FALSE}
ggplot(rossdata_short, aes(x = gdp, y = polity, 
                 color = oil, shape = factor(oecd))) +
  geom_jitter(height = .5,  size = 3, alpha = .4) + 
  theme_bw() 
```



```{}



```

```{r, echo=FALSE}
hi_oil <- rossdata_short %>% filter(oil>10) 

ggplot(rossdata_short, aes(x = gdp, y = polity, 
                 colour = factor(oecd), size = (oil))) +
  geom_jitter(height = .2, alpha = .2) + 
  theme_bw() 
```





It seems like the higher the GDP the more democratic countries are, except if you are a high oil producer or a non-OECD. Let's explore these relationships using a regression.

```{r}
model1 <- lm(polity ~ oil, data = rossdata_short)
summary(model1)
```

Let's now include controls for GDP per capita and OECD membership

```{r}
model2<-lm(polity ~ gdp + oil + oecd, data = rossdata_short)
summary(model2)
```

3. *Interpreting a multivariate model*
    -  Which has a larger impact on the level of democracy: oil dependence or OECD membership?
    -  Which has a larger impact on the level of democracy: oil dependence or GDP per capital?

Recall the OECD membership clustering? Let's try an interaction

```{r}
model3 <- lm(polity ~ gdp + oil*oecd, data = rossdata_short)
summary(model3)
```


4. *Interpreting a multivariate model with an interaction*
    - How would you interpret these results?



## Visualiazing Regression Results

`broom` has three main functions, all of which return data frames (not lists, numeric vectors, or other types of object). `glance` returns a data frame with a single row summary of the model:

```{r}
glance(model2)
```

`tidy` returns a data frame with a row for each coefficient estimate:

```{r}
tidy(model2)
```

`augment` returns the original data frame used in the model with additional columns for fitted values, the standard errors of those fitted values, residuals, etc.

```{r}
head(augment(model2))
```

How about a coefficient plot, roppeladder... etc.

```{r}
ggplot(tidy(model2) %>% filter(term != "(Intercept)"), 
       aes(x = term, y = estimate, 
           ymin = estimate - 2 * std.error, 
           ymax = estimate + 2 * std.error)) + 
  geom_pointrange() + 
  coord_flip()

```

We can also use `coefplot` 
```{r, warning=FALSE, eval = FALSE}
coefplot(model2, coefficients = c("oecd", "oil", "gdp"))
```


5. *Coefficient Plots*
    -  What is wrong with this plot?
    -  Is it useful?
    -  Why? Why not?


## Regression tables

Several packages (`stargazer`, `texreg`, `apsrtable`) are useful for creating publication type regression tables. `stargazer` and `texreg` are the most complete package. Both allow output to either LaTeX or HTML tables for many types of statistical models. We'll use *stargazer* here:

```{r results = 'asis', warning=FALSE, message=FALSE}
library(stargazer)
stargazer(model1, model2, model3, type = "html")
```



## Predicted Values


We are going to use `predict` to get predicted values. We first have to set up a `newdata` 
```{r}
xnew <- data.frame(
  gdp = 5.9, oil = 0, oecd = 1)

predict(model2, newdata = xnew, interval = "confidence")

xnew2 <- data.frame(gdp = 5.9, oil = 0, oecd = 0)

predict(model2, newdata = xnew2, interval = "confidence")
```

What is  this really doing?
```{r}
model2
names(model2)

pe2<-model2$coefficients
pe2
1*pe2[1] + 5.9*pe2[2] + 0*pe2[3] + 1*pe2[4]


1*pe2[1] + 5.9*pe2[2] + 0*pe2[3] + 0*pe2[4]

```

### Going Beyond a simple prediction

We can create a matrix of hypothetical data to obtain predictions for a range of values:


```{r}
xnew <- expand.grid(gdp = seq(4,11),
                    oil = 0,
                    oecd = 1)

xnew <- expand.grid(gdp = seq(4,11),
                    oil = mean(rossdata_short$oil),
                    oecd = mean(rossdata_short$oecd))
```



Now we feed this new data into `predict`
```{r}
pred.res <- predict(model2, newdata = xnew, interval="confidence")

```

### Ploting Predicted Values

To plot these predicted values we have to create a data frame containing both the predicted values generated by predict and the data used to generate those values 
```{r}
mod2_pred_df <- cbind(xnew, pred.res)

mod2_pred_df
```

We  now have a data-frame that can easily be taken by `ggplot`
```{r}
ggplot(mod2_pred_df, aes(x = gdp, y = fit,
                         ymin = lwr, ymax = upr)) +
  geom_line() +
  geom_ribbon(alpha = 0.2)+
  labs(y = "Democracy", x = "GDP")+
  theme_bw()
```

What about oil?

```{r}
xnew <- expand.grid(gdp = mean(rossdata_short$gdp),
                    oil = seq(0,11),
                    oecd = mean(rossdata_short$oecd))

pred.res <- predict(model2, newdata = xnew, interval = "confidence")

mod2_pred_df <- cbind(xnew, pred.res)


ggplot(mod2_pred_df, aes(x = oil, y = fit,
                         ymin = lwr, ymax = upr)) +
  geom_line() +
  geom_ribbon(alpha = 0.2)+
  labs(y = "Democracy", x = "Oil")+
  theme_bw()
```


### Predicting Interactions

Let's get predictions from the model with the interaction.

```{r}
summary(model3)

xnew <- expand.grid(gdp = mean(rossdata_short$gdp),
                    oil = seq(0,11),
                    oecd = c(0,1))



pred.res <- predict(model3, newdata=xnew, interval="confidence")

mod3_pred_df <- cbind(xnew, pred.res)

```


Now we can ggplot it
```{r}
ggplot(mod3_pred_df, aes(x = oil , y = fit, 
                         ymin = lwr, ymax = upr,
                         color = factor(oecd),
                         fill = factor(oecd))) +
  geom_line() +
  geom_ribbon( alpha = 0.7) +
  labs(title = "Predicted Democracy by GDP and OECD membership",
       y = "Predicted Value of Democracy",
       x = "Oil") + 
  scale_fill_discrete(name = "OECD",
                      labels = c("Non Member","Member")) +
  scale_color_discrete(name = "OECD",
                      labels = c("Non Member","Member")) +
  theme_bw()
```

***

<div class="alert alert-danger" role="alert">
Science should be open! Here at Cornell and everywhere, this lab is released under a 
[Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 

</div>



